Regime shifts and transformations in social-ecological systems: 

### 2025-09-30 review_paper/manuscript.Rmd
proof-read version to co-authors

```{r setup, include=FALSE}
library(tidyverse)
library(here)

dat <- googlesheets4::read_sheet(
    "https://docs.google.com/spreadsheets/d/1KmEF9qwiIn--WBDQULL_Plsu5Bw4RJrksnOatGFCQgU/edit?gid=0#gid=0", 
    sheet = 1, skip = 1, col_types = "ccccciiiiiiiiiiiiii"
) |> janitor::clean_names()

dat |> 
    rename(affiliation = 3, author_order = 6) |> 
    arrange(author_order, surname) |> 
    glue::glue_data("<<name>> <<surname>>$^{<<affiliation>>}$",
                    .open = "<<", .close = ">>") |> 
    str_flatten_comma(last = ", and ")

aff <- googlesheets4::read_sheet(
    "https://docs.google.com/spreadsheets/d/1KmEF9qwiIn--WBDQULL_Plsu5Bw4RJrksnOatGFCQgU/edit?gid=0#gid=0", 
    sheet = 2, skip = 0, col_types = "cc"
) |> janitor::clean_names()

glue::glue_data(aff, '    | \\tiny $^{<<code>>}$<<affiliation>>',
                .open = "<<", .close = ">>")

@@

In particular, dimensions that are central to one framework are at risk of becoming oversimplified in the other.
For example, agency —the capacity of actors to act upon their own choices — is central within transformations research [@olsson2006shooting; @moore2014studying; @benessaiah2021] but less prominent within regime shifts studies, with some exceptions in economics [@li2024].
Conversely, the role of feedbacks or time scales are central in explaining regime shifts [@nyström2012; @folke2004; @rocha2018], but not always assessed in transformations research [@olsson2010building], with some exceptions [@enfors2013].

This lack of dialogue has prevented the emergence of a consensus in terminology, leading to oversimplification, and separate jargon denoting similar characteristics.
For example, the concept of positive tipping points emphasizes desirable transformative changes such as the adoption of electric vehicles [@lenton2025; @lenton2023global], but often omits an analysis of equity or which societal groups will benefit or suffer in this transition, and such changes implies for social-ecological sustainability [@pereira2024equity; @pereira2025].
In contrast, issues related to justice and power have been studied for three decades in transformations research [@pereira2024equity; @blythe2018] highlighting that transformations always involve shifting power relations and entail socially differentiated risks and benefits; and can have unintended consequences [@blythe2018].

@@

\begin{figure*}[h!]
\centering
\includegraphics[width =1\textwidth]{fig1_ARER}
\caption{\textbf{Regime shifts and transformation research.} The potential landscape has inspired heuristic models of regime shifts (A) and transformations (B). A cartoon of the cusp bifurcation for forest and savanna shows that the same system can show semi-linear, abrupt but reversible, and discontinuous change with hysteresis (C). The top 30 papers on regime shifts ranked by relevance are organized on a timeline for regime shifts (D) and transformations (F). There is a lack of cross-fertilization between both strands of literature, with an annual proportion of references lower than 5\% and declining since 2020 (E), Figure \ref{fig:cross-ref} dissagregates the proportion per year. The idea of potential attractors is common to both, but there are differences on how they are studied. This represent an opportunity for cross-fertilization.} 
\label{fig:fig1}
\end{figure*}

@@

\begin{textbox}[h]
\subsection{Seeds of Good Anthropocene}

@@

@@

\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{S\arabic{table}}
\setcounter{table}{0}
\setcounter{figure}{0}
\begin{figure*}[ht]
\centering
\includegraphics[width=1\textwidth]{combined-figs}
\caption{\textbf{Lack of cross-fertilization} There has been low cross-referencing between the literature of regime shifts and transformations. Top panels present the proportion of references per year from one literature to another. Data was sourced from OpenAlex with 53380 papers on regime shifts and 97300 papers on sustainability transformations, accessed in September 2025. Data used comes from OpenAlex, an open search engine for scientific literature. Relevance in OpenAlex is computed as the semantic similarity between the search term (e.g. "sustainability transformations") and the text in the title and abstract, weighed by the number of citations.} 
\label{fig:cross-ref}
\end{figure*}

>>>>

```{r setup, include=FALSE, echo=FALSE}
# library(tidyverse)
# library(here)
# 
# dat <- googlesheets4::read_sheet(
#     "https://docs.google.com/spreadsheets/d/1KmEF9qwiIn--WBDQULL_Plsu5Bw4RJrksnOatGFCQgU/edit?gid=0#gid=0", 
#     sheet = 1, skip = 1, col_types = "ccccciiiiiiiiiiiiii"
# ) |> janitor::clean_names()
# 
# dat |> 
#     rename(affiliation = 3, author_order = 6) |> 
#     arrange(author_order, surname) |> 
#     glue::glue_data("<<name>> <<surname>>$^{<<affiliation>>}$",
#                     .open = "<<", .close = ">>") |> 
#     str_flatten_comma(last = ", and ")
# 
# aff <- googlesheets4::read_sheet(
#     "https://docs.google.com/spreadsheets/d/1KmEF9qwiIn--WBDQULL_Plsu5Bw4RJrksnOatGFCQgU/edit?gid=0#gid=0", 
#     sheet = 2, skip = 0, col_types = "cc"
# ) |> janitor::clean_names()
# 
# glue::glue_data(aff, '    | \\tiny $^{<<code>>}$<<affiliation>>',
#                 .open = "<<", .close = ">>")

@@

In particular, dimensions that are central to one framework are at risk of becoming oversimplified in the other.
For example, agency —the capacity of actors to act upon their own choices — is central within transformations research [@olsson2006shooting; @moore2014studying; @benessaiah2021] but less prominent within regime shifts studies, with some exceptions in economics [@li2024].
Conversely, the role of feedbacks or time scales are central in explaining regime shifts [@nyström2012; @folke2004; @rocha2018], but not always assessed in transformations research [@olsson2010building], with some exceptions [@enfors2013]. This lack of dialogue has prevented the emergence of a consensus in terminology, leading to oversimplification, and separate jargon denoting similar characteristics.
For example, the concept of positive tipping points emphasizes desirable transformative changes such as the adoption of electric vehicles [@lenton2025; @lenton2023global], but often omits an analysis of equity or which societal groups will benefit or suffer in this transition, and such changes implies for social-ecological sustainability [@pereira2024equity; @pereira2025].
In contrast, issues related to justice and power have been studied for three decades in transformations research [@pereira2024equity; @blythe2018] highlighting that transformations always involve shifting power relations and entail socially differentiated risks and benefits; and can have unintended consequences [@blythe2018].

@@

\begin{figure*}[ht]
\centering
\includegraphics[width =1\textwidth]{fig1_ARER}
\caption{\textbf{Regime shifts and transformation research.} The potential landscape has inspired heuristic models of regime shifts (A) and transformations (B), while regime shifts focus on plausible states of the system, transformations focus on imagining futures (C). The top 30 papers on regime shifts ranked by relevance are organized on a timeline for regime shifts (D) and transformations (F). Figure \ref{fig:top_papers} shows their bibliometric details. There is a lack of cross-fertilization between both strands of literature, with an annual proportion of references lower than 5\% and declining since 2020 (E), Figure \ref{fig:cross-ref} dissagregates the proportion per year. The idea of potential attractors is common to both, but there are differences on how they are studied. This represent an opportunity for cross-fertilization.} 
\label{fig:fig1}
\end{figure*}

@@

\begin{textbox}[h]
\subsection{Seeds of Good Anthropocenes}

@@

\begin{figure*}[ht]
\centering
\includegraphics[width =1\textwidth]{fig2_ARER}
\caption{\textbf{Empirical evidence} Both regime shifts and transformations used concepts from bifurcation theory as inspiration. A cusp manifold (A) shows different types of transitions using forest and savanna as example, depending on parameter values on one of the drivers (co-dim1) the transition can be smooth (B), abrupt but without hysteresis (C), or discontinuous with hysteresis (a fold bifurcation in D); but along the second driver (co-dim2) the transition is continuous (a fork bifurcation in E). Over 3000 cases of regime shifts are documented on the regime shifts database (F), while over 400 cases of potential transformations are documented on the seeds of good anthropocenes project which has informed the latest IPBES assessment (G). Academic production in number of academic papers is shown for regime shifts (H) and transformations (I) confirm the geographic bias in both areas.} 
\label{fig:fig2}
\end{figure*}

@@

\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{S\arabic{table}}
\setcounter{table}{0}
\setcounter{figure}{0}

\begin{figure*}[ht]
\centering
\includegraphics[width=1\textwidth]{top_papers}
\caption{\textbf{Top 30 papers} Data was sourced from OpenAlex with 53380 papers on regime shifts and 97300 papers on sustainability transformations, accessed in September 2025. OpenAlex is an open search engine for scientific literature. Relevance is computed as the semantic similarity between the search term (e.g. "sustainability transformations") and the text in the title and abstract, weighed by the number of citations.} 
\label{fig:top_papers}
\end{figure*}

\begin{figure*}[ht]
\centering
\includegraphics[width=1\textwidth]{combined-figs}
\caption{\textbf{Lack of cross-fertilization} There has been low cross-referencing between the literature of regime shifts and transformations. Top panels present the proportion of references per year from one literature to another.} 
\label{fig:cross-ref}
\end{figure*}
